---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Spectroscopy datasets often consist of high-dimensional measurements across different wavelengths or wavenumbers, structured in wide-format tables where each column represents a sample and each row corresponds to a spectral position. The `{tidyspec}` package provides a tidyverse-friendly toolkit designed to streamline the processing, visualization, and analysis of such spectral data in R.

By building on top of the `{dplyr}`, `{ggplot2}`, and `{tidyr}` ecosystems, `{tidyspec}` simplifies common workflows such as baseline correction, unit conversion, normalization, filtering, and principal component analysis (PCA). It enforces consistent treatment of spectral dimensions using a customizable reference column (typically wavenumber or wavelength), which can be easily set using `set_spec_wn()`.

This vignette demonstrates a typical usage of the package, from importing and visualizing spectral data to performing PCA and interpreting results. The functions included are designed to make spectral analysis in R more transparent, reproducible, and user-friendly for both beginners and advanced users.

## Instalation

```{r, eval=FALSE}
remotes::install_github("marceelrf/tidyspec")
```

## Package overview

The package organizes utilities into 6 families, all prefixed with `spec_`:

**1. Transformation**  
  `spec_abs2trans()`: Convert absorbance to transmittance.
  `spec_trans2abs()`: Convert transmittance to absorbance.

**2. Normalization**  
  `spec_norm_01()`: Normalize spectra to range [0, 1].
  `spec_norm_minmax()`: Normalize to a custom range.
  `spec_norm_var()`: Scale to standard deviation = 1.

**3. Baseline Correction**  
  `spec_blc_rollingBall()`: Correct baseline using rolling ball algorithm.
  `spec_blc_irls()`: Correct using Iterative Restricted Least Squares.
  `spec_bl_*()`: Return baseline vectors (e.g., `spec_bl_rollingBall`).

**4. Smoothing**  
  `spec_smooth_avg()`: Smooth with moving average.
  `spec_smooth_sga()`: Smooth using Savitzky-Golay.

**5. Derivative**  
  `spec_diff()`: Compute spectral derivatives.

**6. Preview & I/O**  
  `spec_smartplot()`: Static preview of spectra.
  `spec_smartplotly()`: Interactive preview (`{Plotly}`).
  `spec_read()`: Import from .csv, .txt, .xlsx, etc.

```{r,warning=FALSE,message=FALSE}
library(tidyspec)
library(tidyverse)
```


## Data

The `CoHAspec` dataset is a spectral table in wide format, where:

Rows: Wavenumbers (in cm⁻¹).

Columns: Samples (CoHA01, CoHA025, CoHA05, CoHA100).

Values: Absorbance/intensity measurements.

Here’s a preview of the data:

```{r}
CoHAspec
```

### Wavenumber Handling

The function `set_spec_wn` simplifies the use of functions by globally defining the column that contains the wave numbers. User can check the wavenumber column with `check_wn_col`.

```{r}
set_spec_wn("Wavenumber")
check_wn_col()
```

## Visualize the data

Scientists who work with spectroscopy data always want to visualize it after each operation. That's why we created the `spec_smartplot` and `spec_smartplotly` functions to help users.

`spec_smartplot` creates a static visualization of the spectra, while `spec_smartplotly` creates an interactive visualization, allowing the user to navigate through the values of the different data.

```{r,out.width = "100%",fig.alt="Static plot", fig.dpi=300}
spec_smartplot(CoHAspec)
```

```{r,out.width = "100%",fig.alt="Interactive plot", fig.dpi=200}
spec_smartplotly(CoHAspec,geom = "line")
```

These functions simplify the process by automating it, but we invite users to make their own figures for presentations and publications. As part of the efforts to create a community of `tidyspec` users, we plan to create a `ggplot2` extension for high-level spectroscopy graphics. For now, you can produce `ggplot2` graphs by pivoting the data:

```{r, out.width="100%", fig.align="center", fig.height=2, fig.dpi=300, fig.alt="Custom plots", fig.cap="Custom plots"}
CoHAspec %>% 
  tidyr::pivot_longer(cols = -Wavenumber,
                      names_to = "spectrums",
                      values_to = "absorbance") %>% 
  ggplot(aes(x = Wavenumber, y = absorbance, col = spectrums)) +
  geom_line() #<------ Customize your data from here
```


## Convert to transmittance (and back to absorbance)

The functions `spec_abs2trans` and `spec_trans2abs` are designed to easily convert the spectras.

```{r}
CoHAspec %>% 
    spec_abs2trans() %>%
    spec_smartplot()
```

```{r}
CoHAspec %>% 
    spec_abs2trans() %>%
    spec_trans2abs() %>%
    spec_smartplot()
```

## Select spectra

`Tidyspec` is designed to be easily integrated with the entire `tidyverse` ecosystem. Then the user can easily choose which spectra to keep/discard using `dplyr::select`.

```{r}
CoHAspec %>% 
    dplyr::select(Wavenumber,CoHA100) %>%
    tidyspec::spec_smartplot()
```

However, since `tidyspec` needs the column containing the wave numbers and the user could end up messing up and not keeping it during operations, we decided to develop `spec_select` which simplifies the process.

```{r}
CoHAspec %>% 
    tidyspec::spec_select(CoHA100) %>%
    tidyspec::spec_smartplot()
```

## Filter spectra

Similarly to the spectrum selection process, filtering regions of the spectrum can be done using `dplyr::filter`.

```{r}
CoHAspec_filt <- 
    CoHAspec %>% 
    tidyspec::spec_select(CoHA100) %>%
    dplyr::filter(Wavenumber > 1000,
                Wavenumber < 1950)
```

`spec_filter` simplifies the process of filtering the desired regions.

```{r, out.width = "100%",fig.alt="With `spec_select`", fig.dpi=600}
CoHAspec_filt <- 
    CoHAspec %>% 
    spec_select(CoHA100) %>%
    spec_filter(wn_min = 1000,
                wn_max = 1950)

spec_smartplot(CoHAspec_filt, geom = "line")
```

## Smoothing the data

### Moving averages method

```{r}
CoHAspec_filt %>% 
  spec_smooth_avg() %>%
  spec_smartplot(geom = "line")
```

### Savitz-Golay method

```{r}
CoHAspec_filt %>% 
  spec_smooth_sga() %>%
  spec_smartplot(geom = "line")
```

## Derivatives

```{r}
CoHAspec_filt %>% 
    tidyspec::spec_smooth_sga() %>%
    spec_diff(degree = 2) %>%
    tidyspec::spec_smartplot(geom = "line")
```

## Baseline correction

### Rolling ball method

```{r, out.width="100%", fig.dpi=600}
CoHAspec_filt %>% 
    spec_smooth_sga() %>%
    spec_blc_rollingBall(wn_col = "Wavenumber",
                         Wn_min = 1030,
                         Wn_max = 1285,
                         ws = 10, wm = 50) %>%
    tidyspec::spec_smartplot(geom = "line")
```

### Looking to the baseline

```{r, out.width="100%", fig.dpi=600}
CoHAspec_filt %>% 
    spec_smooth_sga() %>%
    spec_bl_rollingBall(wn_col = "Wavenumber",
                         Wn_min = 1030,
                         Wn_max = 1285,
                         ws = 10, wm = 50) %>% 
  spec_smartplot()
```


```{r, out.width="100%", fig.dpi=600}
bl <- CoHAspec_filt %>% 
    spec_smooth_sga() %>%
    spec_bl_rollingBall(wn_col = "Wavenumber",
                         Wn_min = 1030,
                         Wn_max = 1285,
                         ws = 10, wm = 50)

CoHAspec_filt %>%
  spec_smooth_sga() %>%
  spec_filter(wn_min = 1030, wn_max = 1285) %>% 
  left_join(bl) %>% 
  spec_smartplot(geom = "line")
```

## Scaling the spectra

```{r}
CoHAspec_region <- CoHAspec  %>%
    dplyr::filter(Wavenumber > 1300,
                Wavenumber < 1950) 

CoHAspec_region %>% 
    spec_smartplot(geom = "line")
```

### Normalize between 0 and 1

```{r}
CoHAspec_region %>%
    spec_norm_01() %>% 
    spec_smartplot(geom = "line")
```

### Normalize setting the min and max values

```{r}
CoHAspec_region %>%
    spec_norm_minmax(min = 1, max = 2) %>% 
    spec_smartplot(geom = "line")
```

### Normalize to mean = 0 and variance = 1

```{r}
CoHAspec_region %>%
    spec_norm_var() %>% 
    spec_smartplot(geom = "line")
```

## Integration with tidyverse

### Case 1: Choosing the best baseline parameters

```{r}
dados_1030_1285 <- CoHAspec_filt %>% 
    tidyspec::spec_smooth_sga() %>%
    dplyr::filter(Wavenumber <= 1285, Wavenumber >= 1030)

params <- tidyr::crossing(ws_val = c(2,4,6,8,10,12),
                          wm_val = c(10, 25, 40, 50))
params 
```

```{r}
df <- params %>%
    dplyr::mutate(spectra = list(dados_1030_1285)) %>%
    dplyr::mutate(spectra_blc = purrr::pmap(list(spectra, ws_val, wm_val),
                             \(x, y, z ) spec_blc_rollingBall(.data = x, ws = y, wm = z))) %>%
    dplyr::mutate(title = paste0("ws = ", ws_val," , wm = ", wm_val )) %>%
    dplyr::mutate(plot = purrr::map2(spectra_blc, title, ~ spec_smartplot(
            .data = .x,
            wn_col = "Wavenumber",
            geom = "line") + labs(title = .y)
            ))
```

```{r}
library(gridExtra)

grid.arrange(grobs = df$plot[1:4], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[5:8], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[9:12], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[13:16], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[17:20], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[21:24], nrow = 2, ncol = 2)
```

