---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

```{r,warning=FALSE}
library(tidyspec)
library(tidyverse)
```


## Data

```{r}
CoHAspec
```

```{r}
set_spec_wn("Wavenumber")
check_wn_col()
```

## Visualize the data

```{r}
spec_smartplot(CoHAspec)
```


## Convert to transmittance (and back to absorbance)

```{r}
CoHAspec %>% 
    spec_abs2trans() %>%
    spec_smartplot()
```

## Select spectra

```{r}
CoHAspec %>% 
    dplyr::select(Wavenumber,CoHA100) %>%
    tidyspec::spec_smartplot()
```

```{r}
CoHAspec %>% 
    tidyspec::spec_select(CoHA100) %>%
    tidyspec::spec_smartplot()
```

## Filter spectra

```{r}
CoHAspec_filt <- 
    CoHAspec %>% 
    tidyspec::spec_select(CoHA100) %>%
    dplyr::filter(Wavenumber > 1000,
                Wavenumber < 1950)
```

```{r}
CoHAspec_filt <- 
    CoHAspec %>% 
    spec_select(CoHA100) %>%
    spec_filter(wn_min = 1000,
                wn_max = 1950)

spec_smartplot(CoHAspec_filt)
```

## Smoothing the data

### Moving averages method

```{r}
CoHAspec_filt %>% 
  spec_smooth_avg() %>%
  spec_smartplot(geom = "line")
```

### Savitz-Golay method

```{r}
CoHAspec_filt %>% 
  spec_smooth_sga() %>%
  spec_smartplot(geom = "line")
```

## Derivatives

```{r}
CoHAspec_filt %>% 
    tidyspec::spec_smooth_sga() %>%
    spec_diff(degree = 2) %>%
    tidyspec::spec_smartplot(geom = "line")
```

## Baseline correction

```{r}
CoHAspec_filt %>% 
    spec_smooth_sga() %>%
    spec_blc_rollingBall(wn_col = "Wavenumber",
                         Wn_min = 1030,
                         Wn_max = 1285,
                         ws = 10, wm = 50) %>%
    tidyspec::spec_smartplot(geom = "line")
```

## Scaling the spectra

```{r}
CoHAspec_region <- CoHAspec  %>%
    dplyr::filter(Wavenumber > 1300,
                Wavenumber < 1950) 

CoHAspec_region %>% 
    spec_smartplot(geom = "line")
```

### Normalize between 0 and 1

```{r}
CoHAspec_region %>%
    spec_norm_01() %>% 
    spec_smartplot(geom = "line")
```

### Normalize setting the min and max values

```{r}
CoHAspec_region %>%
    spec_norm_minmax(min = 1, max = 2) %>% 
    spec_smartplot(geom = "line")
```

### Normalize to mean = 0 and variance = 1

```{r}
CoHAspec_region %>%
    spec_norm_var() %>% 
    spec_smartplot(geom = "line")
```

## Integration with tidyverse

### Case 1: Choosing the best baseline parameters

```{r}
dados_1030_1285 <- CoHAspec_filt %>% 
    tidyspec::spec_smooth_sga() %>%
    dplyr::filter(Wavenumber <= 1285, Wavenumber >= 1030)

params <- tidyr::crossing(ws_val = c(2,4,6,8,10,12),
                          wm_val = c(10, 25, 40, 50))
params 
```

```{r}
df <- params %>%
    dplyr::mutate(spectra = list(dados_1030_1285)) %>%
    dplyr::mutate(spectra_blc = purrr::pmap(list(spectra, ws_val, wm_val),
                             \(x, y, z ) spec_blc_rollingBall(.data = x, ws = y, wm = z))) %>%
    dplyr::mutate(title = paste0("ws = ", ws_val," , wm = ", wm_val )) %>%
    dplyr::mutate(plot = purrr::map2(spectra_blc, title, ~ spec_smartplot(
            .data = .x,
            wn_col = "Wavenumber",
            geom = "line") + labs(title = .y)
            ))
```

```{r}
library(gridExtra)

grid.arrange(grobs = df$plot[1:4], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[5:8], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[9:12], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[13:16], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[17:20], nrow = 2, ncol = 2)
grid.arrange(grobs = df$plot[21:24], nrow = 2, ncol = 2)
```

