---
title: "FTIR preprocessing - Alginate + gelatin example"
output:
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
---

## Package

```{r}
library(tidyverse)
library(readxl)
library(viridis)
library(baseline)
library(patchwork)
library(plotly)
library(glue)
devtools::load_all()
```


## Import

Importing the data can be done using packages such as `readr`, `readxl`, or the `base` itself. It is important to keep in mind that the `tidyspec` package works using the tidy-data concept.

```{r}
data <- readxl::read_xlsx(path = here::here("AlgGel/dados.xlsx"))


glimpse(data)
```


## Visualize

The `spec_smartplot` and `spec_smartplotly` were created to present a quick visualization of the spectra based on ggplot and plotly, respectively.

```{r,out.width="100%"}
data %>% spec_smartplot()
```

```{r,out.width="100%"}
data %>% spec_smartplotly()
```


## Transform

```{r,out.width="100%"}
data %>% spec_abs2trans() %>% spec_smartplot()
```

## First region of interest:  $1490 cm^{-1} <=$ Wavenumber $<=  1850 cm^{-1}$

### Filter the data

The function `filter` from the `dplyr` package can be used to filter the wavenumber column.

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850)
```

### Smooth the data

The function `spec_smooth_sga` uses the Savitzky-Golay algorithm to smooth the data.

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21)
```

```{r,out.width="100%"}
p1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smartplot() +
  ggtitle(label = "Before")

p2 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_smartplot() +
  ggtitle(label = "After")

(p1 + p2) + plot_layout(guides = "collect")
```

The function `spec_smooth_avg` also can be used to smooth.

### Baseline correction

`spec_blc_rollingBall` function uses the rolling Ball algorithm to perform the baseline correction.

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31)
```

```{r,out.width="100%"}
p1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>%
  spec_smartplot() +
  ggtitle(label = "Before")

p2 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_smartplot() +
  ggtitle(label = "After")

(p1 + p2) + plot_layout(guides = "collect")
```


### Normalize the data

`spec_norm_01` normalize the data to the range of 0 to 1. The maximun value of ALL samples it's used to perform the normalization.

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_01()
```

```{r,out.width="100%"}
p1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_smartplot() +
  ggtitle(label = "Before")

p2 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_01() %>% 
  spec_smartplot() +
  ggtitle(label = "After")

(p1 + p2) + plot_layout(guides = "collect")
```

`spec_norm_minmax` will normalize to a custom range but different from `spec_norm_01` each sample will have it's maximun value set as the maximum value selected.

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_minmax(min = 0,max = 1)
```

```{r,out.width="100%"}
p1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_smartplot() +
  ggtitle(label = "Before")

p2 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_minmax(min = 0,max = 1) %>% 
  spec_smartplot() +
  ggtitle(label = "After")

(p1 + p2) + plot_layout(guides = "collect")
```

`spec_norm_var` can be used to normalize the date to have a standard deviation of one. 

```{r}
data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_var() %>% 
  spec_norm_01() 
```

```{r,out.width="100%"}
p1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_smartplot() +
  ggtitle(label = "Before")

p2 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_var() %>% 
  spec_norm_01() %>% 
  spec_smartplot() +
  ggtitle(label = "After")

(p1 + p2) + plot_layout(guides = "collect")
```

## Exporting the data

You can use functions from the `readr` or `writexl` packages to export your data. 

```{r}

data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_01() %>% 
  #spec_smartplot(xmax = 1785)
  write_csv(file = here::here("reg1_norm.csv"))
```

## View results

```{r}
reg1 <- data %>% 
  dplyr::filter(Wn >= 1490, Wn <= 1850) %>% 
  spec_smooth_sga(window = 21) %>% 
  spec_blc_rollingBall(Wn_min = 1490,Wn_max = 1850,wm = 60,ws = 31) %>% 
  spec_norm_01()
```

```{r,out.width="100%"}
reg1 %>% 
  pivot_longer(cols = -Wn,names_to = "Hidrogel",values_to = "Abs") %>% 
  ggplot(aes(x = Wn, y = Abs, color = Hidrogel)) +
  geom_line(size = 1.4) +
  scale_x_reverse() +
  scale_color_manual(values = c("#800303", "#06147D", "#824EC2", "#054505", "#A36310")) +
  theme_bw()
```

